Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-08-24 GHI_enh_07_Aerosols.tex
Commit 2024-06-06 20:17
Commit 2024-06-09 22:17
Commit 2024-08-24 06:14

\title{Enhancement of SDR in Thessaloniki}
\author{Natsis Athanasios\footnote{Laboratory of Atmospheric Physics, AUTH, \href{mailto:natsisphysicist@gmail.com}{\nolinkurl{natsisphysicist@gmail.com}}} \and Alkiviadis Bais\footnote{Laboratory of Atmospheric Physics, AUTH}}
\date{2024-06-06}

@@

\begin{verbatim}
Registered S3 method overwritten by 'ggpmisc':
  method                  from   
  as.character.polynomial polynom
\end{verbatim}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-1)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-2)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-3)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-4)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-5)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-6)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-7)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-8)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-9)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-10)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-11)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-12)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-13)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-14)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-15)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-4-16)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-1)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-2)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-3)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-4)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-5)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-6)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-7)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-8)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-9)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-10)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-11)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-12)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-13)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-14)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-15)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-16)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-17)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-18)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-19)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-20)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-21)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-22)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-23)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-24)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-25)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-26)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-27)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-28)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-29)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-30)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-31)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-32)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-33)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-34)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-35)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-36)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-37)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-38)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-39)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-40)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-41)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-42)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-43)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-44)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-45)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-46)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-47)
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-48)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-49)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-50)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-51)
\end{figure}

@@

\caption{ - empty caption - }(\#fig:unnamed-chunk-5-52)
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{GHI_enh_07_Aerosols_files/figure-latex/P_CS_change-1}

@@

\caption{ - empty caption - }(\#fig:P_CS_change)
\end{figure}

>>>>

\title{Enhancement of SDR in Thessaloniki}
\author{Natsis Athanasios\footnote{Laboratory of Atmospheric Physics, AUTH, \href{mailto:natsisphysicist@gmail.com}{\nolinkurl{natsisphysicist@gmail.com}}} \and Alkiviadis Bais\footnote{Laboratory of Atmospheric Physics, AUTH}}
\date{2024-08-23}

@@

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-1}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-2}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-3}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-4}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-5}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-6}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-7}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-8}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-9}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-10}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-11}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-12}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-13}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-14}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-15}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-4-16}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-1}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-2}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-3}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-4}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-5}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-6}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-7}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-8}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-9}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-10}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-11}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-12}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-13}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-14}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-15}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-16}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-17}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-18}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-19}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-20}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-21}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-22}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-23}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-24}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-25}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-26}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-27}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-28}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-29}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-30}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-31}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-32}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-33}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-34}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-35}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-36}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-37}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-38}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-39}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-40}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-41}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-42}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-43}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-44}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-45}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-46}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-47}
\end{figure}
\begin{figure}[H]

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-48}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-49}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-50}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-51}
\end{figure}

@@

\caption{ - empty caption - }\label{fig:unnamed-chunk-5-52}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{GHI_enh_07_Aerosols_files/figure-latex/P-CS-change-1}

@@

\caption{ - empty caption - }\label{fig:P-CS-change}
\end{figure}

### 2024-08-24 article/article.tex
Commit 2024-08-24 06:14

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OIR). The OIR is defined as the
irradiance difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}. We have tested both methods by using an
appropriate relative threshold and/or an additional constant offset. The
initial results showed that we can detect a big portion of the actual CE
events. However, by inspecting the daily plots of irradiance it became
evident that changes in atmospheric conditions introduced numerous false
positive or false negative results. The main reason for these
discrepancies is the variability of the effects of aerosols and water
vapor which were not taken into account in the two simple methods. To
produce a more representative reference we included the effects of these
factors using a radiative transfer model (RTM). The applied methodology
is discussed in section\nobreakspace{}\ref{rtmcs}.

@@

This is the criterion of our CE identification. The constant terms were
determined through the implementation of an empirical method with manual
inspection of the CE identification results on selected days of the
whole dataset. We tested seven sets of days with different
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of CE events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of CE events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed, we also used images from the
sky-camera to further aid the decision of the manual inspection.

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance is not known and can
only be approximated. However, this method was proven capable in
detecting all major CE events. Where some CE events with very low OIR
may be not detected, these are few with small over-irradiance, and it is
unlikely that will affect significantly our results.

@@

Following the application of the above discussed methodology to the
entire dataset (\(6.1\) million of one-minute GHI measurements),
\(1.76\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007, corresponding to OIR of
\(345.9\,\text{W}/\text{m}^2\). The strongest CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016. In the
following sections we are discussing the long-term trends and
variability of the CE events as well as of the corresponding OIR and
excess irradiation.

@@

One aspect of this study is to investigate the time evolution of the CE
events by analyzing the GHI measurements at Thessaloniki. Cloud
enhancements can be influenced by different factors, such as the
geometry, size and optical thickness of clouds, their height in the
atmosphere and local weather regimes
\citep{Mol2023, Veerman2022, Gristey2022, Tzoumanikas2016}. Some of
these factors are related to changes in climate; hence it would be
reasonable to expect contributing to the frequency of occurrence of CE
events over Thessaloniki, as well as to the average OIR and excess
irradiation. The long-term trends were calculated using a first-order
autoregressive model with the `maximum likelihood' fitting method
\citep{Gardner1980, Jones1980}, by implementing the function `arima'
from the library `stats' of the R programming language \citep{RCT2023}.
All trends are reported together with their \(2\sigma\) error.

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OIR and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.9 \pm 67.8\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OIR
is negligible (\(+0.1\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OIR for the entire period is
\(+39.9\pm 2.7\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The distribution of the number of CE events as a function of OIR is
shown in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an
inverse relationship between the frequency of CE events and OIR with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favoring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OIR is
below the long-term average of \(39.9\,\text{W}/\text{m}^2\), while
about \(8.1\,\%\) of the events correspond to OIR larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OIR
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

\caption{Relative frequency distribution of CE events in Thessaloniki for the period 1993 -- 2023 as a function of OIR.}\label{fig:ovir-distribution}
\end{figure}

@@

The relation between the duration and the mean OIR of the groups has
also been studied (Figure\nobreakspace{}\ref{fig:group-2d}). Evidently,
events of high excess irradiation have small duration and vice versa.
The vast majority of grouped events are associated with small excess
irradiation (e.g., \(<5\,\text{kJ}/\text{m}^2\)) and small duration
(e.g., \(<5\,\text{min}\)) while groups with strong excess irradiation
and long duration are very rare. Similar results of this relation have
been reported by \citet{Zhang2018} in a study using a far higher
sampling rate (\(100\,\text{Hz}\)) than ours.

@@

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}) shows that in
rare cases the OIR can exceed the TSI even by more than
\(400\,\text{W}/\text{m}^2\), while in \(75\,\%\) of the cases the OIR
is below \(200\,\text{W}/\text{m}^2\). The OIR for the most frequent
ECEs ranges between \(140\) and \(180\,\text{W}/\text{m}^2\). These
findings are in accordance with the results of \citet{Vamvakas2020}, the
only difference being that the OIR values reported for are higher than
those for Thessaloniki.

@@

We have observed ECE cases exceeding the TSI by
\(400\,\text{W}/\text{m}^{2}\) with \(75\,\%\) of the cases under
\(200\,\text{W}/\text{m}^{2}\). The climatological characteristics of
the ECE events showed that the most active months are spread in half of
the year and particularly in the periods March -- June and September --
October. The magnitude of the ECE events identified in Thessaloniki
events does not exceed the values reported for sites with more
favourable conditions for the phenomenon \citep[e.g.,][]{Cordero2023}.
Some of the characteristics of CE and ECE events we analyzed have strong
similarities with the results of \citet{Vamvakas2020} for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude
of OIR.

>>>>

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OI). The OI is defined as the irradiance
difference of the measured one-minute GHI from the
\(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere. First,
we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}. We have tested both methods by using an
appropriate relative threshold and/or an additional constant offset. The
initial results showed that we can detect a big portion of the actual CE
events. However, by inspecting the daily plots of irradiance it became
evident that changes in atmospheric conditions introduced numerous false
positive or false negative results. The main reason for these
discrepancies is the variability of the effects of aerosols and water
vapor which were not taken into account in the two simple methods. To
produce a more representative reference we included the effects of these
factors using a radiative transfer model (RTM). The applied methodology
is discussed in section\nobreakspace{}\ref{rtmcs}.

@@

This is the criterion of our CE identification. The constant terms were
determined through the implementation of an empirical method with manual
inspection of the CE identification results on selected days of the
whole dataset. We tested seven sets of days with different
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest OI CE events, (b) the
largest daily total OI, (c) absence of clouds (by implementing a clear
sky identification algorithm as discussed in \citet{Natsis2023}), (d)
absence of clouds and absence of CE events, (e) with at least \(60\,\%\)
of the day length without clouds and presence of CE events, (h) randomly
selected days, and (i) manually selected days. For the latter case and
where needed, we also used images from the sky-camera to further aid the
decision of the manual inspection.

The definition of the CE events with this method has a degree of
subjectivity, since the actual clear sky irradiance is not known and can
only be approximated. However, this method was proven capable in
detecting all major CE events. Where some CE events with very low OI may
be not detected, these are few with small over-irradiance, and it is
unlikely that will affect significantly our results.

@@

Following the application of the above discussed methodology to the
entire dataset (\(6.1\) million of one-minute GHI measurements),
\(1.76\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007, corresponding to OI of
\(345.9\,\text{W}/\text{m}^2\). The strongest CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016. In the
following sections we are discussing the long-term trends and
variability of the CE events as well as of the corresponding OI and
excess irradiation.

@@

One aspect of this study is to investigate the time evolution of the CE
events by analyzing the GHI measurements at Thessaloniki. Cloud
enhancements can be influenced by different factors, such as the
geometry, size and optical thickness of clouds, their height in the
atmosphere and local weather regimes
\citep{Mol2023, Veerman2022, Gristey2022, Tzoumanikas2016}. Some of
these factors are related to changes in climate; hence it would be
reasonable to expect contributing to the frequency of occurrence of CE
events over Thessaloniki, as well as to the average OI and excess
irradiation. The long-term trends were calculated using a first-order
autoregressive model with the `maximum likelihood' fitting method
\citep{Gardner1980, Jones1980}, by implementing the function `arima'
from the library `stats' of the R programming language \citep{RCT2023}.
All trends are reported together with their \(2\sigma\) error.

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.9 \pm 67.8\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OI
is negligible (\(+0.1\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OI for the entire period is
\(+39.9\pm 2.7\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The distribution of the number of CE events as a function of OI is shown
in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an inverse
relationship between the frequency of CE events and OI with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favoring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OI is
below the long-term average of \(39.9\,\text{W}/\text{m}^2\), while
about \(8.1\,\%\) of the events correspond to OI larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OI
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

\caption{Relative frequency distribution of CE events in Thessaloniki for the period 1993 -- 2023 as a function of OI.}\label{fig:ovir-distribution}
\end{figure}

@@

The relation between the duration and the mean OI of the groups has also
been studied (Figure\nobreakspace{}\ref{fig:group-2d}). Evidently,
events of high excess irradiation have small duration and vice versa.
The vast majority of grouped events are associated with small excess
irradiation (e.g., \(<5\,\text{kJ}/\text{m}^2\)) and small duration
(e.g., \(<5\,\text{min}\)) while groups with strong excess irradiation
and long duration are very rare. Similar results of this relation have
been reported by \citet{Zhang2018} in a study using a far higher
sampling rate (\(100\,\text{Hz}\)) than ours.

@@

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}) shows that in
rare cases the OI can exceed the TSI even by more than
\(400\,\text{W}/\text{m}^2\), while in \(75\,\%\) of the cases the OI is
below \(200\,\text{W}/\text{m}^2\). The OI for the most frequent ECEs
ranges between \(140\) and \(180\,\text{W}/\text{m}^2\). These findings
are in accordance with the results of \citet{Vamvakas2020}, the only
difference being that the OI values reported for are higher than those
for Thessaloniki.

@@

We have observed ECE cases exceeding the TSI by
\(400\,\text{W}/\text{m}^{2}\) with \(75\,\%\) of the cases under
\(200\,\text{W}/\text{m}^{2}\). The climatological characteristics of
the ECE events showed that the most active months are spread in half of
the year and particularly in the periods March -- June and September --
October. The magnitude of the ECE events identified in Thessaloniki
events does not exceed the values reported for sites with more
favourable conditions for the phenomenon \citep[e.g.,][]{Cordero2023}.
Some of the characteristics of CE and ECE events we analyzed have strong
similarities with the results of \citet{Vamvakas2020} for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude
of OI.
